---
title: "Introduction to scDiffCom"
description: >
  Learn how to use scDiffCom to perform differential intercellular communication
  analysis.
author: "Cyril Lagger, Eugen Ursu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to scDiffCom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview and terminology

The goal of scDiffCom is to investigate how cell-cell interactions detected
in scRNA-seq data change between two biological conditions. You can also
simply use it in "detection-mode" if you are not interested in comparing 
conditions.

In addition to this vignette, we also recommend reading our (manuscript
in preparation) that explains the details of the statistical methods used 
by the package. Some important terminology includes:

* LRI: a known ligand-receptor interaction of the
form `Apoe:Ldlr` (simple) or `Col3a1:Itgb1_Itga2` (complex/heteromeric)
* CCI (cell-cell interaction): a communication signal of the form
`(B cell, T cell; Apoe:Ldlr)` where `B cell` is the emitter cell type expressing
the ligand `Apoe` and `T cell` is the receiver cell type expressing the receptor
`Ldlr`.

**Note: all the toy-model results below do not convey any meaningful biology.**

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```

## Standard differential analysis

### Setup a Seurat object

scDiffCom uses a scRNA-seq [Seurat](https://satijalab.org/seurat/) object as
input. Here, we use a toy-model object (available as part of our package) that
has been down-sampled. It contains cells divided in 5 cell types as well as two
age groups ("young" and "old") on which differential analysis will be
performed.

```{r setup}
library(Seurat)
library(scDiffCom)
library(data.table)

# Load the toy-model object
seurat_object <- scDiffCom::seurat_sample_tms_liver
seurat_object

# Cell-type annotations must be present beforehand in the object meta.data
head(seurat_object[["cell_type"]])

# The two conditions of interest must also be specified in the object meta.data
table(seurat_sample_tms_liver[["age_group"]])
```

### (Optional) Investigate the ligand-receptor interactions

scDiffcom infers cell type to cell type communication patterns from the
expression of genes known to be involved in ligand-receptor interactions
(LRIs). The package contains it own internal databases of curated LRIs (for
human, mouse and rat), retrieved from previous studies.

You do not need to call the LRI databases explicitly when performing an
analysis. However, you can still access and explore them as follows:
```{r}
# Load, e.g., the mouse database
data(LRI_mouse)

# Display the data.table of LRIs (more information available in other columns)
LRI_mouse$LRI_curated[, c("LRI")]

# Display the data.table of GO Terms attached to each LRI
LRI_mouse$LRI_curated_GO

# Information about GO terms (more information available in other columns)
data(gene_ontology_level)
gene_ontology_level[, c("NAME", "LEVEL")]

```

### Run default differential analysis

All results are obtained by simply calling `run_interaction_analysis`. The
detection and differential analyses rely on permutation tests. By default,
1000 iterations are performed as this is sufficient for a quick explanatory
analysis. In general, we would recommend using 10'000 permutations. Parallel
computing can be easily enabled by loading the `future` package and setting the
`plan` accordingly.

```{r}
# Load the future package (optional)
library(future)
plan(sequential) # sequentially in the current R process, equivalent to do nothing
#plan(multisession, workers = 4) # background R sessions
#plan(multicore, workers = 4) # forked R processes, not Windows/not RStudio

# Run differential analysis with default parameters
scdiffcom_object <- run_interaction_analysis(
  seurat_object = seurat_object,
  LRI_species = "mouse",
  seurat_celltype_id = "cell_type",
  seurat_condition_id = list(
    column_name = "age_group",
    cond1_name = "YOUNG",
    cond2_name = "OLD"
  )
)
```

The output of `run_interaction_analysis` is an S4 object of class
`scDiffCom`:
```{r}
scdiffcom_object
```

Note: we recommend accessing the slots of the object with the provided
accessors.

### Explore the detected CCIs

The slot `cci_table_detected` is a data.table that contains
all biologically relevant CCIs (one per row) with columns providing useful
information such as how they are regulated between the two conditions
(here `young` and `old`):
```{r}
# Retrieve and display all detected CCIs
CCI_detected <- GetTableCCI(scdiffcom_object, "detected", simplified = TRUE)
CCI_detected[, c("CCI", "REGULATION")]

# Number of CCIs per regulation type
table(CCI_detected$REGULATION)
```

Detected CCIs can be explored either one by one by looking at the data.table,
more globally based on our over-representation analysis (see below) or from 
graphs such as this volcano plot:
```{r,out.width="100%"}
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)

ggplot(
  CCI_detected,
  aes(
     x = LOGFC,
     y = -log10(BH_P_VALUE_DE + 1E-2),
     colour = REGULATION
  )
) + geom_point(
) + scale_colour_manual(
  values = c("UP" = "red", "DOWN" = "blue", "FLAT" = "green", "NSC" = "grey")
) + xlab(
  "log(FC)"
) + ylab(
  "-log10(Adj. p-value)"
)
```


### Explore ORA results

The slot `ora_table` contains a list of data.tables giving the results
of the over-representation analysis performed on predefined categories:
```{r,, out.width="100%"}
# Retrieve the ORA results
ORA_results <- GetTableORA(scdiffcom_object, categories = "all", simplified = TRUE)

# Categories available
names(ORA_results)

# Explore, e.g., the top 10 LRIs up-regulated based on their ORA score
ORA_results$LRI[, c("VALUE", "ORA_SCORE_UP", "OR_UP", "BH_P_VALUE_UP")][order(-ORA_SCORE_UP)][1:10]

# Plot the most over-represented up-regulated LRIs
# note that PlotORA returns a ggplot object that you can further optimize (e.g. here to place the legend)
PlotORA(
  object = scdiffcom_object,
  category = "LRI",
  regulation = "UP"
) + theme(
  legend.position = c(0.85, 0.4),
  legend.key.size = unit(0.4, "cm")
)
```

A summary of the changes in intercellular communication is given by
the following graph that displays the over-represented cell types and cell-type
pairs:
```{r}
if (!require("visNetwork")) install.packages("visNetwork")
if (!require("igraph")) install.packages("igraph")
if (!require("kableExtra")) install.packages("kableExtra")
if (!require("RColorBrewer")) install.packages("RColorBrewer")
library(visNetwork)
library(igraph)
library(kableExtra)
library(RColorBrewer)

BuildNetwork(
  object = scdiffcom_object
)
```

## Additional functionalities

### Modifying default parameters

`run_interaction_analysis` depends on several parameters that can be modified.
We encourage to carefully read the documentation (`?run_interaction_analysis`).
For convenience, the input parameters are stored in the returned object in the
slot `parameters`:
```{r}
# display the first three parameters
head(GetParameters(scdiffcom_object), 3)
```

Among those parameters are four filtering thresholds that can be modified to
recompute the results without having to perform the time-consuming 
permutation analyses again. This works because we store an intermediate
table of all hypothetic CCIs in the slot `cci_table_raw`. The disadvantage is
that storing this data.table can make the object heavy depending on the
dataset. You can either decide to keep this table allowing you to quickly update
your results in case you want to modify some filtering parameters or to erase
this table to save space.


```{r, error=TRUE}
# create a new object with a different log fold change threshold
scdiffcom_object_stringent_logfc <- FilterCCI(
  object = scdiffcom_object,
  new_threshold_logfc = log(2)
)

# create a copy of the original object, but without the heavy intermediate table
scdiffcom_object_light <- EraseRawCCI(scdiffcom_object)

# the light object cannot be updated anymore
FilterCCI(
  object = scdiffcom_object_light,
  new_threshold_logfc = log(2)
)
```

### Performing ORA on custom categories

By default, ORA is performed on a set of categories that are always present for
any dataset analyzed (LRI, GO Terms, etc). However, you might want to find
over-represented results for a category only relevant to your dataset. This is 
possible to achieve as long as you are able to build a relationship between
the terms of your new category and those of a default category. As an example,
we show here how we can perform ORA on cell-type families:

```{r}
# Create a data.table that associates cell-type families to cell types
# Here we use the default category EMITTER_CELLTYPE (RECEIVER_CELLTYPE would also work)
cell_families_relation <- data.table(
  EMITTER_CELLTYPE = c(
    "B cell",
    "T cell",
    "endothelial cell of hepatic sinusoid",
    "hepatocyte",
    "myeloid leukocyte"
  ),
  EMITTER_CELLFAMILY = c(
    "leukocyte",
    "leukocyte",
    "endothelial cell",
    "epithelial cell",
    "leukocyte"
  )
)

# Run ORA with the cell type families as extra annotation
scdiffcom_object <- RunORA(
  object = scdiffcom_object,
  extra_annotations = list(
    cell_families_relation
  ),
  overwrite = FALSE
)
```

### Running a detection analysis without differential analysis

If you are not interested in comparing how CCIs change between two conditions,
you can also use scDiffCom in "detection-mode" only:
```{r}
# not specifying the conditions to compare will perform a detection-only analysis
scdiffcom_detection_only <- run_interaction_analysis(
  seurat_object = seurat_object,
  LRI_species = "mouse",
  seurat_celltype_id = "cell_type",
  seurat_condition_id = NULL
)
```

More downstream analysis tools to explore and visualize such data will be
released in the future.
