---
title: "scDiffCom: GO Terms"
description: >
  Check GO term annotation strategies and evaluate within scDiffCom.
author: "Cyril Lagger, Eugen Ursu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GO Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The objectives of this vignette are:

1. Show some of the details on how the GO Term annotation takes place in `scDiffCom`.
2. Investigate the properties of GO Term merging modes.

Let's start by presenting the relevant data structures used within `scDiffCom`.

## GO Terms data structures in `scDiffCom`

First, there is the gene ontology `data.table` that describes the location of each GO term in the GO ontology graph. In particular, note the integer column `LEVEL` that is a proxy for the distance to the GO term in the GO ontology graph from the "roots" (biological process, molecular function, cellular component).

```{r,error=FALSE,message=FALSE,warning=FALSE,results='hide'}
library(knitr)
library(glue)
library(data.table)
library(ggplot2)
library(easyGgplot2)
library(scDiffCom)

data(gene_ontology_level)
head(gene_ontology_level)
```

Second, there is another `data.table` that connects LRIs to GO terms. For example, for mice, there are two ways of accessing it:

1. Compute
2. Precomputed - internal `scDiffCom` data: `scDiffCom::LRI_mouse$LRI_curated_GO`, computed with (1)

```{r,echo=FALSE,error=FALSE,message=FALSE,warning=FALSE,results='hide'}
## 1. Compute
GO_interactions_genes = scDiffCom:::get_GO_interactions("mouse", scDiffCom::LRI_mouse$LRI_curated, only_genes_annotations = TRUE)

## 2. Precomputed
# scDiffCom::LRI_mouse$LRI_curated_GO

GO_interactions_genes
```

So internally, `get_GO_interactions` performs the assignment of GO Terms to LRI. We allow for several modes of merging the GO terms for the ligands and receptors into a set of GO terms associated with a LRI. Below we compute the `LRI_curated_GO` for each such mode. In what follows, we will compare them.

```{r,include=FALSE}
LR_MERGING_MODES = c(
    "ancestors_intersection", "ancestors_union",
    "ancestors_ligands", "ancestors_receptors",
    "intersection", "union", "ligands", "receptors"
  )

LRI_curated_GO_list = lapply(LR_MERGING_MODES, function(mode) {
  GO = scDiffCom:::get_GO_interactions(
    "mouse", 
    scDiffCom::LRI_mouse$LRI_curated, 
    only_genes_annotations = FALSE,
    LR_merging_mode = mode,
    return_genes_go_ancestors = FALSE)
  GO = GO[!is.na(GO_NAME)]
})
names(LRI_curated_GO_list) = LR_MERGING_MODES
```

Let's see how different are the assignments of GO Terms to LRI as a function of merging modes. Firstly, we check the distribution across LRI of the maximum level of the assigned GO Terms. In other words, for each LRI, we extract the levels of each of the assigned GO Terms and take the maximum over this set.

```{r,fig.width=15,fig.height=5,warning=FALSE}

plots_max = list()
plots_median = list()
for (i in 1:length(LRI_curated_GO_list)) {

  GO = LRI_curated_GO_list[[i]]
  merge_mode = names(LRI_curated_GO_list)[i]
  
  GO = merge(GO, gene_ontology_level, by.x="GO_ID", by.y="ID", all.x=TRUE)
  GO = GO[!is.na(GO$LEVEL)]

  GO_max_level = GO[, list(max_level = max(LEVEL)), by="LRI"]
  plot_max = (ggplot(data=GO_max_level, aes(x=max_level)) 
       + geom_histogram(binwidth = 0.5)
       + xlim(0, 18)
       # + scale_x_continuous(breaks = 1:18)
       + ylim(0, 1200)
       + ggtitle(glue::glue("{merge_mode}\nNum GO:LRI={nrow(GO)}\nN={nrow(GO_max_level)}\n"))
  )
  plots_max[[merge_mode]] = plot_max
  
  GO_median_level = GO[, list(median_level = round(median(LEVEL))), by="LRI"]
  # GO_median_level = GO_median_level[!is.na(GO_median_level$median_level), ]
  plot_median = (ggplot(data=GO_median_level, aes(x=median_level)) 
       + geom_histogram(binwidth = 0.5)
       + xlim(0, 13)
       # + scale_x_continuous(breaks = 1:18)
       + ylim(0, 2100)
       + ggtitle(glue::glue("{merge_mode}\nNum GO:LRI={nrow(GO)}\nN={nrow(GO_median_level)}"))
  )
  plots_median[[merge_mode]] = plot_median
  
}
ggplot2.multiplot(plotlist=plots_max, cols=4)
```
Now let's check the median level distributions:

```{r,fig.width=15,fig.height=5,warning=FALSE}
ggplot2.multiplot(plotlist=plots_median, cols=4)
```

## Overall GO term level distribution

We present below the distribution of levels from the GO Terms. This provides a baseline value of what to expect when examining the levels of the relevant GO Terms in other contexts.

```{r}
GO_LEVEL_NO_OTHER = gene_ontology_level[ASPECT != "other"]
(ggplot(data=GO_LEVEL_NO_OTHER, aes(x=LEVEL)) 
       + geom_histogram(binwidth = 0.5)
       + scale_x_continuous(breaks = 1:18)
       + ggtitle(glue::glue("GO Terms level distribution.\nN={nrow(GO_LEVEL_NO_OTHER)}"))
)
```

We compute the median level for reference.

```{r}
median(GO_LEVEL_NO_OTHER$LEVEL)
```
## Exploring different merging modes on `scDiffCom` toy object.

Now we have the LRI GO annotations obtained by the different merging modes. We test them on the toy object available in the package `scDiffCom::seurat_sample_tms_liver`. First we perform the raw `scDiffCom` analysis. Then we perform `ORA` on the results using the different LRI GO annotations and compare the results. Finally, we plot the ORA results from different merging modes.

```{r,fig.width=15,fig.height=5}

params = list(
    LRI_species = "mouse",
    seurat_celltype_id = "cell_type",
    seurat_condition_id = list(
      column_name = "age_group",
      cond1_name = "YOUNG",
      cond2_name = "OLD"
    ),  
    object_name = "scDiffCom_object",
    seurat_assay = "RNA",
    seurat_slot = "data",
    log_scale = FALSE,
    score_type = "geometric_mean",
    threshold_min_cells = 5,
    threshold_pct = 0.1,
    iterations = 1000,
    threshold_quantile_score = 0.2,
    threshold_p_value_specificity = 0.05,
    threshold_p_value_de = 0.05,
    threshold_logfc = log(1.5),  # log(1.5)
    return_distributions = FALSE,
    seed = 42,
    verbose = TRUE
  )
  
results = scDiffCom:::run_internal_raw_analysis(
  seurat_object = scDiffCom::seurat_sample_tms_liver,
  LRI_table = scDiffCom::LRI_mouse$LRI_curated,
  LRI_species = "mouse",
  params = params
)
results = scDiffCom::FilterCCI(results)


results_ora = list()
for (i in 1:length(LRI_curated_GO_list)) {
  GO = LRI_curated_GO_list[[i]]
  merge_mode = names(LRI_curated_GO_list)[i]
  tryCatch(
    {
      results_ora[[merge_mode]] = scDiffCom:::run_ora(
        object = results,
        categories = "GO_TERMS",
        extra_annotations = NULL,
        overwrite = TRUE,
        stringent_or_default = "default",
        stringent_logfc_threshold = NULL,
        verbose = TRUE,
        class_signature = "scDiffCom",
        global = FALSE,
        LRI_curated_GO = GO
      )
    },
    error=function (cond) {
      print(cond)
    }
  )
}

dt_ora = data.table::data.table()
for (i in 1:length(results_ora)) {
  merge_mode = names(results_ora)[i]
  res = results_ora[[merge_mode]]
  tryCatch(
    {
      dt = res@ora_table$GO_TERMS
      dt_m = data.table::melt(
        dt, 
        id.vars = c("VALUE_BIS", "VALUE", "CATEGORY", "LEVEL"), 
        measure.vars = c("BH_P_VALUE_UP", "BH_P_VALUE_DOWN", "BH_P_VALUE_FLAT"), 
        value.name = c("P_VAL")
      )
      dt_m$variable = as.character(dt_m$variable)
      dt_m$TYPE = purrr::map_chr(dt_m$variable, function(s) {strsplit(s, split="_")[[1]][4]})
      dt_m = dt_m[dt_m$P_VAL < 0.05, ]
      dt_m$MERGE_MODE = merge_mode
      
      dt_ora = rbind(dt_ora, dt_m)
    },
    error=function (cond) {
      print(cond)
    }
  )
}

dt_ora = dt_ora[!is.na(LEVEL)]
dt_ora = dt_ora[TYPE %in% c("UP", "DOWN")]

violin_plot = ggplot(data=dt_ora, aes(y=LEVEL, x=MERGE_MODE)) + 
  geom_violin() + 
  scale_x_discrete(labels = c(
    "ancestors_intersection" = glue("ancestors_intersection\nN={table(dt_ora$MERGE_MODE)['ancestors_intersection']}"),
    "intersection" = glue("intersection\nN={table(dt_ora$MERGE_MODE)['intersection']}"),
    "ancestors_union" = glue("ancestors_union\nN={table(dt_ora$MERGE_MODE)['ancestors_union']}"),
    "union" = glue("union\nN={table(dt_ora$MERGE_MODE)['union']}"),
    "ancestors_ligands" = glue("ancestors_ligands\nN={table(dt_ora$MERGE_MODE)['ancestors_ligands']}"),
    "ligands" = glue("ligands\nN={table(dt_ora$MERGE_MODE)['ligands']}"),
    "ancestors_receptors" = glue("ancestors_receptors\nN={table(dt_ora$MERGE_MODE)['ancestors_receptors']}"),
    "receptors" = glue("receptors\nN={table(dt_ora$MERGE_MODE)['receptors']}")
  )) +
  xlab("GO Terms merging modes") +
  ylab("Level") +
  ggtitle("Distribution of GO Term levels by merging modes.") +
  theme(axis.text.x = element_text(size=12, angle=0))
violin_plot
```

