---
title: "scDiffCom: GO Annotation"
description: >
  Check GO term annotation strategies and evaluate within scDiffCom.
author: "Cyril Lagger, Eugen Ursu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GO Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The objective of this vignette is to:

1. showcase how the GO Term annotation takes place in scDiffCom.
2. parameter for choosing the GO Term annotation strategy.
3. evaluate the performance of the different strategies.

Let's start by presenting the GO term data structures used within scDiffCom.

## GO Terms data structures in `scDiffCom`

First, there is the gene ontology `data.table` that describes the location of each GO term in the GO ontology graph. In particular, note the integer column `LEVEL` that is a proxy for the distance to the GO term in the GO ontology graph from the (pseudo)roots.

```{r}
data(gene_ontology_level)
head(gene_ontology_level)
```

Second, there is another `data.table` that connects LRIs to GO terms. For example, for mice, there are multiple ways of finding it:

1. compute for LRI
2. compute for genes in LRI
3. internal `scDiffCom` data: `scDiffCom::LRI_mouse$LRI_curated_GO`, computed with (1)

```{r}
## Compute
# GO_interactions_genes = scDiffCom:::get_GO_interactions("mouse", scDiffCom::LRI_mouse$LRI_curated, only_genes_annotations = TRUE)
# GO_interactions = scDiffCom:::get_GO_interactions("mouse", scDiffCom::LRI_mouse$LRI_curated, only_genes_annotations = FALSE)
scDiffCom::LRI_mouse$LRI_curated_GO
```

Internally, `get_GO_interactions` computes intersection of ancestors. Let's see alternative strategies.

## 